# Entropy calculations for AA Multiple Sequence Alignments (MSAs).

Author: Andres Posbeyikian 
Date created: 01/04/2024
Last modified: 22/05/2024
 
## 0. Load libraries
```{r}
suppressWarnings({ 
library(entropy)
library(ggseqlogo)
library(Biostrings)
library(ggplot2)
library(openxlsx)
library(readr)
library(scales)
library(dplyr)
}) 
```


## 1. Set working and output directories, load raw alignment file, and set PATH to Clipkit
```{r}
# Set the working directory
wd <- "/Users/posbeyik/Desktop/ZAR1_AF2_Mult/02_Analysis"
setwd(wd)

# Create a directory to save outputs to
output_dir <- file.path(wd, "Entropy_Analysis")

# Check if the directory already exists, and create it if it does not
if (!dir.exists(output_dir)) {
  dir.create(output_dir)
  message("Directory created at: ", output_dir)
} else {
  message("Directory already exists at: ", output_dir)
}

# Load the raw MSA file
Raw_MSA_path <- "/Users/posbeyik/Desktop/ZAR1_AF2_Mult/02_Analysis/ZAR1_sequences.afa"
Raw_MSA <- readAAStringSet(Raw_MSA_path)

#--------------------------------#
# You may need to first open a terminal and run the following, and then restart the R session:
# > which clipkit
# > export PATH="/path/to/clipkit:$PATH" #Replace the path with the result from running the first line.
#--------------------------------#

# Specify the full path to the ClipKIT executable
clipkit_path <- "/Users/posbeyik/opt/anaconda3/bin/clipkit"
```


## 3. Convert the MSA into a DataFrame object
```{r}
# Function to convert a Biostrings/dataframe object to a dataframe and break the sequence into characters

convert_to_dataframe <- function(alignment) {
  # Extract sequences as strings
  sequences <- sapply(alignment, as.character)
  # Split each sequence into individual characters and convert to a matrix
  alignment_matrix <- do.call(rbind, strsplit(sequences, ""))
  # Convert the matrix to a dataframe
  alignment_df <- as.data.frame(alignment_matrix, stringsAsFactors = FALSE)
  # Naming columns as positions
  colnames(alignment_df) <- paste(seq_len(ncol(alignment_df)))
  
  return(alignment_df)
}

Raw_MSA_df <- convert_to_dataframe(Raw_MSA)
```

## 4. Trim the alignment using Clipkit's tool, Gappy
```{r}

# Extracting the basename without extension and the extension
file_base <- tools::file_path_sans_ext(basename(Raw_MSA_path))
file_ext <- tools::file_ext(Raw_MSA_path)

# Constructing the new output filename
output_filename <- sprintf("%s_trimmed.%s", file_base, file_ext) # The ClipKit output will have the 'trimmed' suffix.
output_path <- file.path(dirname(Raw_MSA_path), output_filename)

# Output path for the log
log_filename <- sprintf("%s.log", output_filename) 
log_path <- file.path(dirname(Raw_MSA_path), log_filename)

# Constructing the full ClipKIT command with the new output file specified
args <- c(shQuote(Raw_MSA_path), "-l", "-m gappy", "-o", shQuote(output_path))

# Execute the command
result <- system2(clipkit_path, args, wait = TRUE)

# Check if the command was successful
if (result == 0) {
    cat("ClipKIT ran successfully!\n")
} else {
    cat("ClipKIT encountered an error.\n")
}
```

## 5. Load the trimmed alignment file and save it as a dataframe
```{r}
Trimmed_MSA <- readAAStringSet(output_path)
Trimmed_MSA_df <- convert_to_dataframe(Trimmed_MSA)
```

## 6. Define a function to calculate Entropy values
```{r}
# Function to calculate entropy for multiple columns and return as a dataframe
calculate_entropy_df <- function(dataframe) {
  # Function to calculate entropy for a single column
  calculate_column_entropy <- function(column) {
    # Remove any NA values (if there are any)
    column <- na.omit(column)
    # Calculate frequencies
    freq_table <- table(column)
    # Convert to proportions
    proportions <- freq_table / sum(freq_table)
    # Calculate and return entropy
    entropy.empirical(proportions, unit = "log2")
  }
  
  # Ensure that the dataframe is processed column-wise
  entropy_values <- sapply(dataframe, calculate_column_entropy, simplify = FALSE)
  
  # Flatten the list if necessary
  if (is.list(entropy_values)) {
    entropy_values <- unlist(entropy_values)
  }
  
  # Create a dataframe with positions and entropy values
  entropy_df <- data.frame(Position = seq_along(entropy_values), Entropy = entropy_values)
  
  return(entropy_df)
}
```

## 7. Calculate the entropy values for the trimmed MSA
```{r}
Trimmed_MSA_entropy <- calculate_entropy_df(Trimmed_MSA_df)

# Save an excel file with the entropy values:
write.xlsx(Trimmed_MSA_entropy, file.path(output_dir,'Trimmed_MSA_Entropy.xlsx'))
```

## 8. Map entropy scores with original positions in untrimmed alignment
```{r}
# The entropy analysis is performed on the trimmed alignment. Therefore, some aminoacids that were in the full MSA have been excluded and do not have an entropy score assigned to them. Positions in the trimmed alignment are not the same as positions in the raw alignment. To correct for this, we need to re-calculate the original positions of the residues.

# We check which columns (aminoacids) in the raw MSA were trimmed, and re-assign the original positions for the entropy values.


# Read Gappy's log file
# Adjust the sep (separator) and header arguments based on the log file's format
log_data <- read_table(log_path, col_names = c("position", "trimmed/kept", "parsimony","gappyness"), col_types = cols())


total_rows <- nrow(log_data)
trim_pos <- 1

#Create an empty list 'trimmed_positions_list'
trimmed_positions_list <- list()

#Start a for loop from 1 to the total number of rows in the dataframe
for (i in 1:total_rows) {
  #Fetch the value from column 'position' of the dataframe at the current index
  position_i <- log_data$position[i]
  
  #Fetch the value from column 'trimmed/kept' of the dataframe at the current index
  status_i <- log_data$`trimmed/kept`[i]

  # Check the value of 'status_i'
  if (status_i == 'trim') {
    # If 'status_i' is 'trim', append NA to 'trimmed_positions_list'
    trimmed_positions_list[[i]] <- NA
  } else if (status_i == 'keep') {
    # If 'status_i' is 'keep', append 'trim_pos' to 'trimmed_positions_list'
    trimmed_positions_list[[i]] <- trim_pos
    # Increment 'trim_pos' by 1
    trim_pos <- trim_pos + 1
  }
}

# Convert list to a vector for appending to dataframe
trimmed_positions_vector <- unlist(trimmed_positions_list)

# Append this vector as a new column named 'trimmed position' to 'log_data'
log_data$`trimmed position` <- trimmed_positions_vector

merged_dataframe <- merge(log_data, Trimmed_MSA_entropy, by.x = "trimmed position", by.y = "Position", all.x = TRUE)

# Exporting Data
# Select the columns we want to export
Raw_Position_Entropy_annotation_df <- merged_dataframe[, c("position", "Entropy")]

#Transpose the dataframe
t_Raw_Position_Entropy_annotation_df <- as.data.frame(t(Raw_Position_Entropy_annotation_df))

# Set the column names of transposed_df to the values of its first row
colnames(t_Raw_Position_Entropy_annotation_df) <- as.character(t_Raw_Position_Entropy_annotation_df[1, ])
# Remove the now redundant first row
t_Raw_Position_Entropy_annotation_df <- t_Raw_Position_Entropy_annotation_df[-1, ]

row_to_append <- t_Raw_Position_Entropy_annotation_df[1, ]  # Select the row to append
Raw_MSA_df <- rbind(Raw_MSA_df, row_to_append)

# Export to a .txt file
write.table(Raw_MSA_df, file.path(output_dir,"Raw_MSA_Entropy.csv"), sep = ",", row.names = TRUE, col.names = TRUE)

```

## 9. Process entropy data for each sequence in the MSA. Make format easier for export into ChimeraX.
```{r}
process_MSA_Data <- function(df) {
  # Initialize a list to collect data for each species
  results_list <- list()
  
  # Iterate through each row, except the row named 'Entropy'
  for (i in 1:nrow(df)) {
    if (rownames(df)[i] != "Entropy") {
      species_name = rownames(df)[i]
      
      # Create a list to store vectors of amino acid and entropy score
      species_data <- list()

      # Iterate through each column
      for (j in 1:ncol(df)) {
        amino_acid <- df[i, j]
        
        # Skip gaps in the alignment (amino acid == "-")
        if (amino_acid == "-") {
          next
        }
        entropy_score <- df['Entropy', j]
        
        # Append the vector (amino_acid, entropy_score) to the species list
        species_data[[length(species_data) + 1]] <- c(amino_acid, as.numeric(entropy_score))
      }

      # Add the list for this species to the results list
      results_list[[species_name]] <- species_data
    }
  }

  # Return the results list
  return(results_list)
}

MSA_species_list <- process_MSA_Data(Raw_MSA_df)
```

## 10. Generating a dataframe for one species of interest
```{r}
# Define the function to create a dataframe for a given species name
create_species_dataframe <- function(species_name, MSA_species_list) {
  # Get the list of tuples for the given species
  species_data <- MSA_species_list[[species_name]]
  
  # Initialize vectors to store position, amino acid identity, and entropy
  positions <- 1:length(species_data)
  amino_acids <- vector("character", length(species_data))
  entropy_scores <- vector("numeric", length(species_data))
  
  # Loop through each position to extract amino acid identity and entropy
  for (i in seq_along(species_data)) {
    amino_acids[i] <- species_data[[i]][1]
    entropy_scores[i] <- species_data[[i]][2]
  }
  
  # Create a dataframe with the collected data
  species_df <- data.frame(
    Position = positions,
    AminoAcid = amino_acids,
    Entropy = entropy_scores
  )
  
  # Convert the 'Entropy' column to numeric (floats), keeping non-numeric as NA
  species_df$Entropy <- suppressWarnings(as.numeric(species_df$Entropy))
  return(species_df)
}

# Example usage
# Assuming MSA_species_list is already defined and populated
species_name <- "AXY05280.1_Nicotiana_benthamiana"
species_df <- create_species_dataframe(species_name, MSA_species_list)

# Print the resulting dataframe
print(species_df)

```

## 11. Chimera-X coloring commands
```{r}

map_entropy_to_hex <- function(df, color_min, color_max) {
  # Ensure the 'Entropy' column exists
  if (!"Entropy" %in% colnames(df)) {
    stop("DataFrame must contain an 'Entropy' column.")
  }
  
  # Calculate min and max of the entropy
  min_entropy <- min(df$Entropy, na.rm = TRUE)
  max_entropy <- max(df$Entropy, na.rm = TRUE)
  
  # Create a colormap including white as the midpoint color
  colormap <- colorRampPalette(c(color_min, "white", color_max))
  
  # Normalize entropy values and apply the colormap
  df$Hex_Code <- sapply(df$Entropy, function(x) {
    normalized <- (x - min_entropy) / (max_entropy - min_entropy)
    colormap(100)[round(normalized * 99) + 1]
  })
  
  return(df)
}

# Define hex colors for min and max
color_min <- '#1d4b8e'  # Lowest Entropy
color_max <- '#db282b'  # Highest Entropy

# Apply the function
hex_species_df <- map_entropy_to_hex(species_df, color_min, color_max)

# We remove all NA values
noNA_hex_species_df <- hex_species_df[!is.na(hex_species_df$Entropy), ]

# Writing the ChimeraX command into a text file
# Assuming your dataframe is called 'data' and the text file is named 'colors.txt'

# Open a connection to a text file
output_file <- file("/Users/posbeyik/Desktop/For AA/Chimera_X_Commands/ChimeraX_commands.txt", open = "wt")


# Modified midway entropy to value of 1.5 / Code by GPT4:

map_entropy_to_hex <- function(df, color_min, color_max) {
  # Ensure the 'Entropy' column exists
  if (!"Entropy" %in% colnames(df)) {
    stop("DataFrame must contain an 'Entropy' column.")
  }
  
  # Calculate min and max of the entropy
  min_entropy <- min(df$Entropy, na.rm = TRUE)
  max_entropy <- max(df$Entropy, na.rm = TRUE)
  
  # Hardcode the midpoint entropy value
  midpoint_entropy <- 1.5
  
  # Check if the midpoint entropy is within the current range
  if (midpoint_entropy < min_entropy || midpoint_entropy > max_entropy) {
    stop("The fixed midpoint entropy of 1.5 is outside the range of data entropy.")
  }
  
  # Calculate lengths of the color map segments
  lower_scale_length <- round(100 * (midpoint_entropy - min_entropy) / (max_entropy - min_entropy))
  upper_scale_length <- 100 - lower_scale_length
  
  # Create color gradients
  lower_colormap <- colorRampPalette(c(color_min, "white"))(lower_scale_length)
  upper_colormap <- colorRampPalette(c("white", color_max))(upper_scale_length)
  
  # Combine the two color maps
  colormap <- c(lower_colormap, upper_colormap)
  
  # Normalize entropy values and apply the colormap
  df$Hex_Code <- sapply(df$Entropy, function(x) {
    if (x <= midpoint_entropy) {
      normalized <- (x - min_entropy) / (midpoint_entropy - min_entropy)
      index <- round(normalized * (lower_scale_length - 1)) + 1
      return(lower_colormap[index])
    } else {
      normalized <- (x - midpoint_entropy) / (max_entropy - midpoint_entropy)
      index <- round(normalized * (upper_scale_length - 1)) + 1
      return(upper_colormap[index])
    }
  })
  
  return(df)
}

# Apply the function
hex_species_df <- map_entropy_to_hex(species_df, color_min, color_max)

# Remove all NA values
noNA_hex_species_df <- hex_species_df[!is.na(hex_species_df$Entropy), ]

# Continue with file output if needed
output_file <- file("/Users/posbeyik/Desktop/For AA/Chimera_X_Commands/ChimeraX_commands.txt", open = "wt")






# Iterate over each row and write the formatted string to the file
apply(noNA_hex_species_df, 1, function(row) {
  position <- row['Position']
  hex_code <- row['Hex_Code']
  line <- paste0("color /a:", position, " ", hex_code)
  writeLines(line, output_file)
})

# Close the connection to the text file
close(output_file)

```

